---
title: "Set 2: FASTQ quality improvement report"
output: 
  pdf_document:
    fig_width: 9
    fig_height: 4
colorlinks: true
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(ShortRead)
library(Biostrings)
library(knitr)

# Show only what is required in the PDF and save R Markdown figures as PNG
knitr::opts_chunk$set(
  fig.path = "figures/",
  dev = "png",
  dpi = 150,
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  purl = FALSE
)

if (!dir.exists("figures")) {
  dir.create("figures", recursive = TRUE)
}

# Plot settings
par(cex.axis = 0.8, cex.lab = 0.9, cex.main = 1.0)

cat("Give the name of .fastq file for analysis, e.g. ex02.fastq: ", file = stderr())

fastqname <- readLines("stdin", n = 1)
fastqname <- trimws(fastqname)

if (fastqname == "") {
  stop("Wrong FASTQ filename or file does not exist.")
}

if (!file.exists(fastqname)) {
  stop(paste("File not found:", fastqname))
}

step_plot <- 5

# Decision from the BEFORE plot:
k_trim <- 50     # trim 50 bases from the end, so the first 200 bases are kept
```

## What this PDF contains

This report includes:

1. A boxplot of per-base Phred quality **before** processing.
2. A boxplot of per-base Phred quality **after** processing.
3. A short explanation of the decisions and how the quality improvement was implemented.

The required quality improvement consists of:

- removing all reads that contain at least one unknown base (`N`)
- trimming the same number of bases from the end of every read

## Read data and compute quality before processing

```{r read-before, include=FALSE}
fastq_raw <- readFastq(fastqname)
seq_raw   <- sread(fastq_raw)
qual_raw  <- as(quality(fastq_raw), "matrix")

n_reads_raw <- length(fastq_raw)

# Readable quality boxplot
quality_boxplot <- function(qual_mat, main, step = 5) {
  idx <- seq(1, ncol(qual_mat), by = step)

  boxplot(
    qual_mat[, idx, drop = FALSE],
    xlab = "Base position",
    ylab = "Phred quality",
    main = main,
    range = 0,         # whiskers show the full quality range
    outline = FALSE,   # hide outliers to make the plot easier to read
    lwd = 0.3,
    lty = 1,
    xaxt = "n"
  )

  axis(1, at = seq_along(idx), labels = idx)
  grid(nx = NA, ny = NULL)
}
```

### Quality before processing

The boxplot below shows per-base Phred quality across reads, displayed every `r step_plot` positions.

**Decision based on the BEFORE plot:**  
The BEFORE chart shows that the quality is very good and stable up to around 190-200 positions. The median is usually around 37-39. After that, there is a clear decline and a larger spread of quality values. The final positions, especially around 230-251, contain many bases with quality around Q20 and below, and the last positions drop even closer to Q15. Therefore, trimming was done by removing about 50 bases from the end, leaving the first 200 bases.

```{r plot-before, fig.cap=paste0(fastqname, " - BEFORE: per-base Phred quality boxplot")}
quality_boxplot(
  qual_raw,
  main = paste0(fastqname, " - BEFORE"),
  step = step_plot
)
```

```{r save-before-png, include=FALSE}
# Save the BEFORE plot as a PNG file for the portfolio website
png("figures/fastq_before.png", width = 1200, height = 700, res = 150)

quality_boxplot(
  qual_raw,
  main = paste0(fastqname, " - BEFORE"),
  step = step_plot
)

invisible(dev.off())
```

## Implementation of quality improvement

### Step 1: filtering

All reads containing at least one unknown base (`N`) are removed.

### Step 2: trimming

A fixed number of bases is trimmed from the end of every read.  
Here, I trim **k = `r k_trim`** bases from the end of each read. The same value is used for all reads and was selected based on the BEFORE quality plot.

## Apply filtering and trimming

```{r process, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
# 1) Remove reads containing N
n_count <- letterFrequency(seq_raw, letters = "N")[, 1]
keep_noN <- (n_count == 0)
fastq_noN <- fastq_raw[keep_noN]

n_removed_N <- sum(!keep_noN)

# 2) Trim a fixed number of bases from the end
end_pos <- width(sread(fastq_noN)) - k_trim
keep_len <- end_pos >= 1

fastq_trim <- narrow(
  fastq_noN[keep_len],
  start = 1,
  end = end_pos[keep_len]
)

n_after_trim <- length(fastq_trim)
qual_after <- as(quality(fastq_trim), "matrix")
```

### Quality after processing

```{r plot-after, fig.cap=paste0(fastqname, " - AFTER: after removing reads with N and trimming a fixed number of bases from the end")}
quality_boxplot(
  qual_after,
  main = paste0(fastqname, " - AFTER (no N, -", k_trim, " bases)"),
  step = step_plot
)
```

```{r save-after-png, include=FALSE}
# Save the AFTER plot as a PNG file for the portfolio website
png("figures/fastq_after.png", width = 1200, height = 700, res = 150)

quality_boxplot(
  qual_after,
  main = paste0(fastqname, " - AFTER (no N, -", k_trim, " bases)"),
  step = step_plot
)

invisible(dev.off())
```

## Short summary

- Input reads: **`r n_reads_raw`**
- Reads removed due to `N`: **`r n_removed_N`**
- Fixed trimming length from the end: **`r k_trim`** bases
- Reads after processing: **`r n_after_trim`**

The plots saved for the portfolio website are:

- `figures/fastq_before.png`
- `figures/fastq_after.png`

---

## Executable R script

The extracted `.R` script accepts a FASTQ file name as a command-line argument or asks for it interactively if it is missing. It removes reads containing at least one `N`, trims the same number of bases from the end of every read, and saves two output files: the processed FASTQ file and the final quality boxplot as a PNG image. The user can choose the trim length, plot step and output compression.

```{r script-set2, echo=FALSE, eval=FALSE, comment="", purl=TRUE}
#!/usr/bin/env Rscript

# Project 4 / Set 2: FASTQ filtering + fixed trimming + final quality plot

suppressPackageStartupMessages({
  library(ShortRead)
  library(Biostrings)
})

input <- function(prompt) {
  cat(prompt)
  flush.console()
  readLines("stdin", n = 1)
}

args <- commandArgs(trailingOnly = TRUE)

# Input FASTQ name, from argument or interactive question
if (length(args) >= 1 && nzchar(args[1])) {
  fastqname <- args[1]
} else {
  fastqname <- input("Enter the name of a FASTQ file: ")
}

if (!file.exists(fastqname)) {
  stop("File not found: ", fastqname)
}

# Trim length, fixed number of bases removed from the end
trim_txt <- if (length(args) >= 2 && nzchar(args[2])) {
  args[2]
} else {
  input("How many bases to trim from the end? (default 50): ")
}

if (!nzchar(trim_txt)) trim_txt <- "50"

TRIM_END <- suppressWarnings(as.integer(trim_txt))

if (is.na(TRIM_END) || TRIM_END < 0) {
  stop("Invalid TRIM_END: ", trim_txt)
}

# Plot step, show every n-th base on the x-axis
step_txt <- if (length(args) >= 3 && nzchar(args[3])) {
  args[3]
} else {
  input("Plot every n-th base position? (default 5): ")
}

if (!nzchar(step_txt)) step_txt <- "5"

STEP <- suppressWarnings(as.integer(step_txt))

if (is.na(STEP) || STEP < 1) {
  stop("Invalid STEP: ", step_txt)
}

# Output compression
comp_txt <- if (length(args) >= 4 && nzchar(args[4])) {
  args[4]
} else {
  input("Compress output FASTQ? [Y/n]: ")
}

comp_txt <- tolower(trimws(comp_txt))
compress_out <- !(comp_txt %in% c("n", "no", "0"))

# Read FASTQ
fastq <- readFastq(fastqname)

# FILTER: remove reads containing at least one N
seq <- sread(fastq)
n_count <- letterFrequency(seq, letters = "N")[, 1]
keep_noN <- (n_count == 0)
filtered <- fastq[keep_noN]

if (length(filtered) == 0) {
  stop("No reads left after filtering. All reads contain N.")
}

# TRIM: fixed number of bases from the end, the same rule for all reads
L <- width(sread(filtered))
minL <- min(L)

if (TRIM_END >= minL) {
  stop(
    "TRIM_END is too large. Minimum read length is ",
    minL,
    " but TRIM_END is ",
    TRIM_END,
    "."
  )
}

endpos <- L - TRIM_END
trimmed <- narrow(filtered, start = 1, end = endpos)

# Output names
base <- tools::file_path_sans_ext(basename(fastqname))

out_fastq <- if (compress_out) {
  paste0(base, "_set2_processed.fastq.gz")
} else {
  paste0(base, "_set2_processed.fastq")
}

out_png <- paste0(base, "_set2_final_quality.png")

# writeFastq fails if the output already exists, so remove old file first
if (file.exists(out_fastq)) {
  file.remove(out_fastq)
}

writeFastq(trimmed, out_fastq, compress = compress_out)

# Final quality boxplot after processing, saved as PNG
qual <- as(quality(trimmed), "matrix")
idx <- seq(1, ncol(qual), by = STEP)

png(out_png, width = 1200, height = 800, res = 150)

par(cex.axis = 0.9, cex.lab = 0.9, cex.main = 1.0)

boxplot(
  qual[, idx, drop = FALSE],
  xlab = "Base position",
  ylab = "Phred quality",
  range = 0,
  outline = FALSE,
  lwd = 0.3,
  lty = 1,
  xaxt = "n",
  main = paste0(basename(fastqname), " - AFTER (filter N + trim ", TRIM_END, ")")
)

axis(1, at = seq_along(idx), labels = idx)
grid(nx = NA, ny = NULL)

invisible(dev.off())

cat("Saved processed FASTQ: ", out_fastq, "\n")
cat("Saved final quality PNG:", out_png, "\n")
cat("Reads total:   ", length(fastq), "\n")
cat("Reads removed: ", sum(!keep_noN), "\n")
cat("Reads kept:    ", length(trimmed), "\n")
cat("Trimmed bases: ", TRIM_END, "\n")
```

```{r auto-purl, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE, purl=FALSE}
# This chunk automatically extracts the embedded script into a standalone .R file when knitting.

Rmdfile <- knitr::current_input()
rmd_norm <- normalizePath(Rmdfile, winslash = "/", mustWork = TRUE)

extractR <- function(input) {
  knitr::purl(input, documentation = 0)
}

ok <- FALSE

if (requireNamespace("callr", quietly = TRUE)) {
  try({
    callr::r(extractR, args = list(rmd_norm))
    ok <- TRUE
  }, silent = TRUE)
}

if (!ok) {
  rscript <- Sys.which("Rscript")

  if (nzchar(rscript)) {
    expr <- sprintf('knitr::purl("%s", documentation = 0)', rmd_norm)
    try(system2(rscript, c("-e", expr)), silent = TRUE)
  }
}
```
